# NYC's Property Valuation and Assessment

Abisola Sajuyigbe

## Introduction

This project, inspired by an interest in financial risk assessments, will explore New York City's Property Valuation and Assessment Data. This chapter displays the exploratory data analysis and visualization of the City's financial evaluations that determine tax revenues. Since property assessments determine tax revenues, identifying gaps between assessed property value and its market value is an important fiscal concern.

This project's goal is to encourage financial data transparency by generating accessible and easy to understand visualizations that encourage public engagement with the city's finances and property valuation. A better New York is one with more informed stakeholders!

This chapter will engage with the following key questions:

1.  How do tax exemption programs influence property values?

2.  What is the relationship between tax classification and market value?

3.  How do tax exemption programs and property market value vary by borough?

4.  How does the valuation gap vary by tax classification and borough?

## Data

```{r}
#| message: false
library(tidyverse)

# -------------------------------SAMPLING----------------------------------
# Base API URL
base_url <- "https://data.cityofnewyork.us/resource/yjxr-fw8i.csv"

# Queries for each borough
boro_queries <- paste0(base_url, "?$where=boro='", 1:5, "'&$limit=10000")

# Pull the data and fix the variable type mistmatches
PVA_stratified <- map_df(boro_queries, ~read_csv(.x, col_types = cols(
  bble = col_character(),   
  boro = col_character(),
  taxclass = col_character(),
  zip = col_character(),
  .default = col_guess() 
)))

# Check distribution and column types
print(dim(PVA_stratified))
PVA_stratified %>% count(boro)



# ----------------------------CLEANING-----------------------------------------

PVA_tidy <- PVA_stratified %>% 
  select(boro, taxclass, fullval, avtot, extot, exmptcl) %>% 
  mutate(across(c(boro, taxclass, exmptcl), as.factor)) %>% 
  mutate(exmpt_code = case_match(exmptcl,
    "VI" ~ "Veterans",
    "X1" ~ "NYC Government",
    "X2" ~ "NY State Government",
    "X3" ~ "Federal Government",
    "X4" ~ "Public Authorities",
    "X5" ~ "Non-Profit/Religious",
    "X6" ~ "Educational",
    "X7" ~ "Benevolent/Charitable",
    "X8" ~ "Hospitals/Health",
    "X9" ~ "Miscellaneous",
    .default = exmptcl
  )) %>%
  
  mutate(taxcl_code = case_match(taxclass,
    "1" ~ "1-3 Unit Residential",
    "1A" ~ "Single-Family Home",
    "1B" ~ "Vacant Land (Residential)",
    "1C" ~ "Condos (in 1-3 unit buildings)",
    "2" ~ "Residential (4+ Units)",
    "2A" ~ "4-6 Unit Buildings",
    "2B" ~ "7-10 Unit Buildings",
    "2C" ~ "2-10 Unit Co-ops/Condos",
    "3" ~ "Utility Property",
    "4" ~ "Commercial & All Other",
    .default = taxclass
  )) %>%
                          mutate(boro_name = case_match(boro,
                            "1" ~ "Manhattan",
                            "2" ~ "Bronx",
                            "3" ~ "Brooklyn",
                            "4" ~ "Queens",
                            "5" ~ "Staten Island"
                          ))


str(PVA_tidy)
```

### Description

The data used in this study is the NYC Property Valuation and Assessment Data, collected by the City government's Department of Finance (DoF), and sourced from [NYC Open Data](https://data.cityofnewyork.us/City-Government/Property-Valuation-and-Assessment-Data/yjxr-fw8i/about_data). It contains property valuation and assessment data for 9.85 million New York City properties, with a collection of 40 variables. The dataset is classified by NYC Open Data as historical, and it is not subject to automated updates.

The initial data export using an API endpoint was limited to 1,000 rows or observations. This form of data extraction created a sampling error, as only observations from the first three listed boroughs were included in the extraction. To ensure the dataset used for analysis is representative of the city, this report uses a stratified sample of 50,000 observations exported via a `SODA2 API` endpoint in `CSV` format, ensuring the most up-to-date figures. This sampling method pulled in data from five distinct batches, one for each borough, using a limit of 10,000 rows per batch. These batches were then aggregated into a comprehensive and balanced dataset of 50,000 rows.

The research conducted in this chapter is targeted, using a "tidied" version of the originally sourced dataset, containing the following variables:

+----------------+------------------------------------------------------------------------------------------------------------------------------+----------------------+
| Variable Name  | Description                                                                                                                  | Data Type            |
+================+==============================================================================================================================+======================+
| **`BORO`**     | Classification of New York City's five boroughs.                                                                             | Factor (Nominal)     |
|                |                                                                                                                              |                      |
|                | -   **1** = Manhattan                                                                                                        |                      |
|                |                                                                                                                              |                      |
|                | -   **2** = Bronx                                                                                                            |                      |
|                |                                                                                                                              |                      |
|                | -   **3** = Brooklyn                                                                                                         |                      |
|                |                                                                                                                              |                      |
|                | -   **4** = Queens                                                                                                           |                      |
|                |                                                                                                                              |                      |
|                | -   **5** = Staten Island                                                                                                    |                      |
+----------------+------------------------------------------------------------------------------------------------------------------------------+----------------------+
| **`TAXCLASS`** | Classification of property determining tax and assessment rates.                                                             | Factor (Ordinal)     |
|                |                                                                                                                              |                      |
|                | -   **1** = 1-3 Unit Residential                                                                                             |                      |
|                |                                                                                                                              |                      |
|                |     -   **1A** = Single-Family Home                                                                                          |                      |
|                |                                                                                                                              |                      |
|                |     -   **1B** = Vacant Land (Residential)                                                                                   |                      |
|                |                                                                                                                              |                      |
|                |     -   **1C** = Condos (in 1-3 unit buildings)                                                                              |                      |
|                |                                                                                                                              |                      |
|                | -   **2** = Residential (4+ Units)                                                                                           |                      |
|                |                                                                                                                              |                      |
|                |     -   **2A** = 4-6 Unit Buildings                                                                                          |                      |
|                |                                                                                                                              |                      |
|                |     -   **2B** = 7-10 Unit Buildings                                                                                         |                      |
|                |                                                                                                                              |                      |
|                |     -   **2C** = 2-10 Unit Co-ops/Condos                                                                                     |                      |
|                |                                                                                                                              |                      |
|                | -   **3** = Utility Property                                                                                                 |                      |
|                |                                                                                                                              |                      |
|                | -   **4** = Commercial & All Other                                                                                           |                      |
+----------------+------------------------------------------------------------------------------------------------------------------------------+----------------------+
| **`FULLVAL`**  | The property's market value.                                                                                                 | Numeric (Continuous) |
+----------------+------------------------------------------------------------------------------------------------------------------------------+----------------------+
| **`AVTOT`**    | The property's assessed value, including the land any structures on it.                                                      | Numeric (Continuous) |
+----------------+------------------------------------------------------------------------------------------------------------------------------+----------------------+
| **`EXTOT`**    | The total amount of the property's assessed value that is not taxable.                                                       | Numeric (Continuous) |
+----------------+------------------------------------------------------------------------------------------------------------------------------+----------------------+
| **`EXMPTCL`**  | Classification for specific property tax exemption programs.                                                                 | Factor (Nominal)     |
|                |                                                                                                                              |                      |
|                | -   **VI**: Veterans exemptions                                                                                              |                      |
|                | -   **X1**: NYC Government Owned (City agencies, parks, public schools).                                                     |                      |
|                | -   **X2**: NY State Government Owned (State offices, authorities).                                                          |                      |
|                | -   **X3**: Federal Government Owned (Post offices, federal courts).                                                         |                      |
|                | -   **X4**: Public Authorities (e.g., MTA, Port Authority, Housing Authority).                                               |                      |
|                | -   **X5**: Private Non-Profit / Public Benefit (Religious, charitable, or educational organizations that are fully exempt). |                      |
|                | -   **X6**: Educational Institutions (Private schools, universities).                                                        |                      |
|                | -   **X7**: Benevolent / Charitable (Non-profit headquarters, foundations).                                                  |                      |
|                | -   **X8**: Hospitals / Health Facilities (Non-profit hospitals, clinics).                                                   |                      |
|                | -   **X9**: Miscellaneous (Specialized non-profits or unique statutory exemptions).                                          |                      |
+----------------+------------------------------------------------------------------------------------------------------------------------------+----------------------+

The data dictionary used to define the variables was incomplete, requiring further outside research on property and real estate terminology and classifications.

### Missing value analysis

```{r}
library(redav)

# Missing values for each variable
colSums(is.na(PVA_tidy)) %>%
  sort(decreasing = TRUE)

# Figure 1. Visualize Aggregated Missing Pattern Plot
library(patchwork)
plot_missing(PVA_tidy, percent = FALSE) &
  theme_classic() &
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 8)
  ) +
  labs(caption = "Figure 1.")
```

**Figure 1.** The Aggregated Missing Pattern Plot indicates that the `exmptcl` and `exmpt_code` variables are the primary sources of missing values in the tidied dataset. Approximately 47,914 rows are missing for these variables, while all other variables remain complete. This missingness pattern is logical and reflective of New York City property tax structures because many properties do not qualify for government or non-profit exemptions. Therefore, they correctly lack a corresponding exemption classification code.

```{r}
# Figure 2. Missingness by Borough and Property Type
missing_data <- PVA_tidy %>%
  group_by(boro_name, taxcl_code) %>%
  summarise(percent_na = mean(is.na(exmptcl)) * 100)

library(ggplot2)

ggplot(missing_data, 
       aes(x = percent_na, y = fct_reorder(taxcl_code, percent_na))) +
  geom_col(fill = "pink2", alpha = 0.6) +
  facet_wrap(~boro_name) +
  theme_classic() +
  labs(title = "Missingness by Borough & Property Type",
       y = "Tax/Property Classification",
       x = "Missing Values (%)",
       caption = "Figure 2.")
```

**Figure 2**. The Missingness by Borough and Property Type Plot indicates that Commercial & All Other properties have the lowest amount of missing data (roughly 80-85%), suggesting that these commercial properties have more recorded tax breaks than residential properties across all five boroughs. Besides that, Manhattan's Single-Family Home bar is missing, reflecting the reality that there are not many single family homes in Manhattan.

## Results

### Tax Exemption Total and Market Values

```{r}
library(scales)
library(ggridges)
  
# Figure 3. 2D Density Plot
ggplot(PVA_tidy, aes(x = extot, y = fullval)) +
  geom_bin2d(bins = 100) + 
  scale_fill_viridis_c(trans = "log10") +
    geom_smooth(method = "lm", se = TRUE, color = "red") +
  scale_x_log10(labels = label_dollar(scale_cut = cut_short_scale())) + 
  scale_y_log10(labels = label_dollar(scale_cut = cut_short_scale())) +
  theme_minimal() +
  labs(title = "Exemption Total vs. Market Value", 
       x = "Exemption Total (Log Scale)",
       y = "Full Property Market Value (Log Scale)",
       caption = "Figure 3.")
```

**Figure 3**. This 2D density plot visualizes the concentration of properties based on their total tax exemption and full market value. Unlike a standard scatterplot, this binned heatmap reveals the volume of data in high-density regions; the brighter yellow bins indicate a significantly higher concentration of properties compared to the darker purple areas. A clear positive correlation is visible, confirming that as property market values increase, the corresponding tax exemption amounts also rise. The high-intensity diagonal band represents a massive concentration of 100% tax-exempt properties, while the vertical concentration identifies a high volume of properties receiving standardized exemption amounts regardless of market value.

Note: Due to the logarithmic scale, this plot only visualizes the subset of properties with a non-zero exemption total.

```{r}
# Figure 4. Ridgeline Plot
PVA_tidy %>%  
   filter(!is.na(exmptcl)) %>%
  filter(extot > 0 & fullval > 0) %>%
   ggplot(aes(x = fullval, y = fct_reorder(exmpt_code, fullval, .fun = median), fill = exmpt_code)) +
   geom_density_ridges(fill = "cyan3", alpha = 0.7) + 
   scale_x_log10(labels = label_dollar(scale_cut = cut_short_scale())) +
   theme_ridges() +
   labs(title = "Market Value Distribution by Exemption Category",
        x = "Full Property Market Value (Log Scale)",
        y = "Exemption Classification Program",
        fill = "Exemption Classification",
       caption = "Figure 4.")
```

**Figure 4**. The ridgeline plot illustrates the association between tax programs and full property market values. The veterans category is clustered toward the bottom left, suggesting that veteran tax programs are associated with properties with the lower market values (likely residential homes). Meanwhile, federal and public authorities, peak towards the right, meaning they have higher valuations. Categories with longer tails reflect the diversity in the property market value of those agencies.

### Tax Classifications and Market Value

```{r}
# Figure 5. Boxplot
PVA_tidy %>%
  ggplot(aes(x = fct_reorder(taxcl_code, fullval, .fun = median), 
             y = fullval, fill = taxcl_code)) +
  geom_boxplot(outlier.alpha = 0.05) + 
  scale_y_log10(labels = label_dollar(scale_cut = cut_short_scale())) +
  theme_minimal() + 
  theme(legend.position = "none") +
  coord_flip(ylim = c(10000, 100000000)) + 
  labs(title = "Focused Market Value Distribution",
       y = "Market Value (Log Scale - Zoomed)",
       x = "Tax Classification",
       caption = "Figure 5.")
```

**Figure 5**. The boxplot indicates that condos and 7-10 unit residential properties have the highest median market value at around \$1M. Meanwhile, utility properties have the lowest median market value at around \$500K.

NOTE: To avoid the compression of boxplots due to outliers that force the IQR to appear as a flat line, this figure provides a "zoomed" view, focusing on the most relevant and common range (e.g., \$10K to \$100M). Overall, this allows for better interpretability of median differences.

```{r}
# Figure 6. Ridgeline Plot
PVA_tidy %>% 
  filter(fullval > 0) %>%
  ggplot(aes(x = fullval, y = fct_reorder(taxcl_code, fullval, .fun = median))) +
  geom_density_ridges(fill = "green3", alpha = 0.7) + 
  scale_x_log10(labels = label_dollar(scale_cut = cut_short_scale())) +
  theme_ridges() +
  labs(title = "Market Value Distribution by Tax Classification",
         x = "Full Property Market Value (Log Scale)",
         y = "Tax/Property Classification",
       fill = "Tax/Property Classification",
       caption = "Figure 6.")
```

**Figure 6**. The ridge-line plot reveals that the distribution for utility properties are multimodal with three distinct peaks and a wide spread. Thus, there are likely three major types of utility properties. Vacant Land (Residential) and Condos (in 1-3 unit buildings) are bimodal with 2 peaks, suggesting two leading types of land properties. Commercial & All Other properties have weak peaks with long tails, and a median market value around \$500-700k, highlighting that the distribution is influenced by outliers.

### Market Value and Tax Exemption by Borough

```{r}
# Figure 7. Cleveland Dot Plot
  boro_stats <- PVA_stratified %>%
                          group_by(boro) %>%
                          summarize(
                            median_market_val = median(fullval, na.rm = TRUE)
                          ) %>%
                          mutate(boro_name = case_match(boro,
                            "1" ~ "Manhattan",
                            "2" ~ "Bronx",
                            "3" ~ "Brooklyn",
                            "4" ~ "Queens",
                            "5" ~ "Staten Island"
                          ))

boro_stats %>%
  ggplot(aes(x = median_market_val, y = fct_reorder(boro_name, median_market_val))) +
  geom_point(size = 4, color = "firebrick") +
  scale_x_log10(labels = label_dollar()) +
  theme_minimal() +
  labs(title = "Median Market Value by Borough",
         x = "Median Market Value", 
         y = "Borough",
       caption = "Figure 7.")
```

**Figure 7**. The Cleveland Dot Plot confirms Manhattan properties have the greatest median market value, in comparison to all the other New York City boroughs. Meanwhile, Queens properties have the lowest median market value in comparison to all the other New York City boroughs.

### Risk by Tax Classification and Borough

```{r}
#---------------------------CALCULATING THE VALUATION GAP---------------------------------
PVA_ratio <- PVA_stratified %>%
  filter(fullval > 0) %>%
  mutate(assessment_ratio = avtot / fullval,
         boro_name = case_match(boro,
            "1" ~ "Manhattan", "2" ~ "Bronx", "3" ~ "Brooklyn",
            "4" ~ "Queens", "5" ~ "Staten Island")) %>%
  mutate(taxcl_code = case_match(taxclass,
    "1" ~ "1-3 Unit Residential",
    "1A" ~ "Single-Family Home",
    "1B" ~ "Vacant Land (Residential)",
    "1C" ~ "Condos (in 1-3 unit buildings)",
    "2" ~ "Residential (4+ Units)",
    "2A" ~ "4-6 Unit Buildings",
    "2B" ~ "7-10 Unit Buildings",
    "2C" ~ "2-10 Unit Co-ops/Condos",
    "3" ~ "Utility Property",
    "4" ~ "Commercial & All Other",
    .default = taxclass
  )) %>%
  filter(assessment_ratio > 0 & assessment_ratio < 1)

# Figure 8. Boxplot
ggplot(PVA_ratio, aes(x = taxcl_code, y = assessment_ratio, fill = taxcl_code)) +
  geom_boxplot(outlier.alpha = 0.1) +
  facet_wrap(~boro_name) +
  geom_hline(yintercept = c(0.06, 0.45), linetype = "dashed", color = "red") +
  scale_y_continuous(labels = scales::label_percent()) +
  coord_cartesian(ylim = c(0, 0.60)) + 
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  labs(title = "Focused Assessment Ratio Distribution",
       x = "Tax Classification",
       y = "Assessment Ratio (%)",
       fill = "Tax Classification",
       caption = "Figure 8.")

```

**Figure 8**. This graph was generated using a calculated assessment ratio, or the portion of a property's market value (Market Value, represented by '`fullval`') that the city decides to tax (Assessed Total, represented by '`avtot`'). The devised formula is $$Assessment\ Ratio = \frac{avtot}{fullval}$$It displays dashed lines at the 6% and 45% Assessment Ratio mark. These represent New York City's target assessment ratio for residential homes and all other properties, respectively. When the assessment ratio is lower than the dashed line, there is a valuation gap, meaning that the property based on its market value is under-taxed. Across all boroughs, the assessment ratios for 1-3 unit residential properties are consistently below the 6% benchmark, suggesting a valuation gap. Additionally, for 4+ unit residential properties, the assessment ratios across all boroughs are consistently below the 45% benchmark, suggesting a valuation gap. It is evident there is model risk in the city's taxation of properties and their market value, which are under-assessed.

## Conclusion

This analysis reveals that New York City's property tax system and valuation assessments are generally consistent with real estate frameworks. However, there are model risks evident in the identified valuation gaps across different tax classifications and boroughs.

This study found that tax exemption programs are associated with property market value. Besides that, tax exemption programs are usually standardized for Veterans, while the highest dollar exemptions are granted to large-scale federal and publicly-owned properties.

Secondly, the valuation gap is the greatest in 4-6 unit buildings, 7-10 unit buildings, and 2-10 unit co-ops or condos, as they consistently fail to meet the 45% assessment ratio benchmark. This is an indication that these building types tend to be under-taxed relative to their current market value.

Nevertheless, some properties, particularly Utility properties are multimodal with high variability. Therefore, it is harder to create a model that will full capture the diversity of NYC's real estate and property portfolio.

This study is limited by its stratified sampling, which only provided 50,000 observations. Additionally, due to the incompleteness in the data dictionary and code book, this project consulted with outside resources to supplement property and real estate knowledge. Nevertheless, the results still emphasize the need for greater transparency and model reforms for New York City's properties.

In the future, this study would like to investigate with a socioeconomic lens, using demographic data to explore the effects of attributes such as age, gender, income, education, and location on property valuation, tax revenues, and home ownership. This would be key to understanding the social science of real estate and potential socioeconomic inequalities. Highlighting these gaps contribute to better policy change.
