# NYPD Arrest Data

By: Jade Amaya

## Introduction

Hi everyone, welcome to my project! This project looks at NYPD Arrest Data (YTD) and examines the relationships between arrests, location, demographic characteristics, and other crime indicators. I use precinct-level aggregated data to analyze how arrest patterns vary across New York City (NYC) and how they relate to complaints, shootings, and race with respect to arrests. I wanted to look at this data set (and related ones) because I have a broader interest in how policing and crime intersect with historical patterns in law enforcement, specifically in the U.S. Some historical context motivating my project is related to slavery patrol systems during the antebellum period that were replaced by modern police systems during the Jim Crow era to reinforce segregation, and recently (as in a few years ago) stop-and-frisk in the late 20th to early 21st century. Amid this context, I use the following datasets to look at statistical relationships that may tell a story as to how arrests are distributed across NYC during this time.

My project is divided into the following three topics, each addressing separate questions:

1. Spatial patterns in arrests

2. The relationship between arrests and crime indicators

3. Racial composition of people arrested across NYC


## Data

### Description

The arrest data was collected by the NYPD, reviewed and extracted by the Office of Management and Planning, then published by NYC’s Open Data team. It was imported, using a csv file, as a dataframe containing 19 columns and 69.3k rows. The data columns spanned information ranging from information about the crime and location to demographics of the person arrested. The data is updated quarterly throughout the year, and most recently during late April. Other datasets that I use in this project are NYPD Complaint data and NYC Shootings, both of which are collected by the NYPD, updated quarterly, and were called into this project using an API. To control for time and location, datasets were filtered to include only  observations from 1/1/26 to 3/31/26 and were grouped and merged by precinct. Some limitations of the dataset are that is does not indicate whether arrested individuals were released, formally charged and indicted, or scheduled for a trial, nor does it clarify whether observations identified repeated offenders. Moreover, some demographic variables contain missing values, and an arrest record by itself does not confirm that an individual committed the alleged offense.

Below are the links to all the datasets and tools used in this project:

[NYPD Arrests](https://data.cityofnewyork.us/Public-Safety/NYPD-Arrest-Data-Year-to-Date-/uip8-fykc/about_data)

[NYPD Complaints](https://data.cityofnewyork.us/Public-Safety/NYPD-Complaint-Data-Current-Year-To-Date-/5uac-w243/about_data)

[NYC Shootings](https://data.cityofnewyork.us/Public-Safety/Shootings-2006-Present-/5ucz-vwe8/about_data)

[NYPP Shape Data (25c)](https://www.nyc.gov/content/planning/pages/resources/datasets/police-precincts)

[NYBB Shape Data (26a)](https://www.nyc.gov/content/planning/pages/resources/datasets/borough-boundaries)

[Color Pallettes Used (Not Populated by R)](https://clrs.cc/)

[The Codebook Bible](https://ap4347.github.io/2102/index.html)

### Missing Value Analysis

```{r}
library(tidyverse)
library(sf)
library(readxl)
library(patchwork)
arrests <- read_csv("data/arrests.csv", na = c("(null)", ""))
```

```{r}
arrests <- arrests |>
  mutate(mdydate = mdy(ARREST_DATE)) |>
  rename(perp_sex = PERP_SEX, age = AGE_GROUP, pd_desc = PD_DESC,
         arrest_key = ARREST_KEY, law_cat_cd = LAW_CAT_CD,
         law_code = LAW_CODE, x_coord_cd = X_COORD_CD, y_coord_cd = Y_COORD_CD,
         ky_cd = KY_CD, jurisdiction_code = JURISDICTION_CODE, 
         ofns_desc = OFNS_DESC, arrest_boro = ARREST_BORO, 
         Precinct = ARREST_PRECINCT, perp_race = PERP_RACE,
         pd_cd = PD_CD, longitude = Longitude, latitude = Latitude) |>
    mutate(arrest_boro = recode(arrest_boro, `M` = "Manhattan", `B` = "Bronx", `K` = "Brooklyn",
                           'Q' = "Queens", 'S' = "Staten Island"))
```

```{r}
#| fig-width: 8
#| warning: false
library(redav)
arrests_new <- arrests |>
  select(starts_with(c("perp_sex", "age", "pd_d", "k", "o", "law_", "pd_c", "arrest_key")))
colnames(arrests_new) <- str_remove_all(colnames(arrests_new), "PE")
plot_missing(arrests_new)
```

The missing value plot shows that most of the rows in the dataset contained no missing values, while the most common missing data patterns involve the perp_sex and age_group variables either by themselves or alongside offense-related variables such as ofns_desc and pd_desc. The remaining 10 columns did not contain missing data and thus were left unlisted.

## Results

### Topic 1: Spatial Patterns in  Arrests

#### Where are arrests concentrated?

```{r}
#|warning: false
library(sf)
nybb <- st_read("nybb_26a/nybb.shp")

nybb <- nybb |>
  mutate(BoroCode = recode(BoroCode, `1` = "Manhattan", `2` = "Bronx", `3` = "Brooklyn",
                           '4' = "Queens", '5' = "Staten Island")) |>
  rename(arrest_boro = BoroCode)
```

```{r}
graph1 <- ggplot(arrests, aes(x = arrest_boro, fill = arrest_boro)) +
  geom_bar() +
  labs(title = "Arrest Counts in New York City by Borough",
       subtitle = "From January 1, 2026 to March 31, 2026",
       y = "count", x = "NYC Borough") +
  theme_minimal()
```

```{r}
points_sf <- arrests |> 
 select(longitude, latitude, arrest_boro) |> 
  na.omit() |> 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) 
graph2 <- ggplot() + 
 geom_sf(data = nybb) +
  geom_sf(data = points_sf, size = .1, alpha = 0.5, aes(color = arrest_boro)) +
  geom_jitter() +
  theme_minimal() + 
  labs(title = "Spatial Distribution of Arrests in New York City",
       subtitle = "From January 1, 2026 to March 31, 2026",
       x = "longitude", y = "latitude") +
  theme(legend.position = "none")
```

##### Concentration by Borough

```{r}
#| fig-length: 5
#| fig-width: 10
graph2 + graph1
```
The graphs compare arrest patterns across NYC's boroughs. The left spatial graph shows the spatial distribution of individual arrest locations using their respective x and y coordinates, while the right bar graph shows the total number of arrests in each borough. Side by side, the plots show that Brooklyn (n = 21111) has the highest numbers of arrests compared to the other boroughs, while Staten Island (n = 2551) has the lowest.

##### Concentration by Precinct

```{r}
#|fig-width: 12

top_precincts_counts <- arrests |>
  count(Precinct, arrest_boro) |>
  rename(count = n) |>
  arrange(desc(count)) |>
  slice_head(n = 10)

ggplot(top_precincts_counts, aes(x = count, y = fct_reorder(as.factor(Precinct), count)
                                 , color = arrest_boro)) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("Bronx" = "#0074D9", "Brooklyn" = "#2ECC40", 
       "Manhattan" = "#85144b", "Queens" = "#FF4136", "Staten Island" = "#FF851B")) +
  geom_text(aes(label = count), color = "black", hjust = -0.3, size = 2.5) +
  labs(y = "Precinct", x = "Number of Arrests", title = "Top 10 Precincts With Highest Number of Arrests",
  subtitle = "Total Number of Arrests from January 1, 2026 to March 31, 2026", color = "Borough") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5, size = 9))
```
The Cleveland dot plot shows the 10 precincts with the highest numbers of arrests between January 1, 2026 to March 31, 2026. The colors of each dot represent the borough to which the precinct belongs. While Brooklyn takes the lead for having the highest amount of arrests, the Bronx accounts for five precincts among the 10. Meanwhile, no precinct from Staten Island placed among the top 10.

```{r}
bottom_precincts_counts <- arrests |>
  count(Precinct) |>
  rename(count = n) |>
  arrange(count) |>
  slice_head(n = 10)

bottom_precincts_counts <- bottom_precincts_counts |>
  left_join(arrests |>
   group_by(Precinct) |>
     summarize(arrest_boro = first(arrest_boro)), by = "Precinct")

ggplot(bottom_precincts_counts, aes(x = count, y = fct_reorder(as.factor(Precinct), count),
                                    color = arrest_boro)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Bronx" = "#0074D9", "Brooklyn" = "#2ECC40", 
        "Manhattan" = "#85144b", "Queens" = "#FF4136", "Staten Island" = "#FF851B")) +
  geom_text(aes(label = count), color = "black", hjust = 2, size = 2.5) +
  labs(y = "Precinct", x = "Number of Arrests", title = "Bottom 10 Precincts With Lowest Number of Arrests",
  subtitle = "Total number of arrests from January 1, 2026 to March 31, 2026", color = "Borough") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5, size = 9))
```
This Cleveland dot plot shows the 10 precincts with the lowest numbers of arrests between January 1, 2026 to March 31, 2026. The color of each point represent the borough to which the precinct belongs. Manhattan contains the precinct with the lowest number of arrests and accounts for four of the ten precincts with the lowest arrest counts. No precinct in the Bronx placed among the bottom 10.

### Topic 2: Relationship Between Arrests and Indicators of Crime Across Precincts

#### What is the association between the number of crime complaints and arrests across precincts?

```{r}
complaints <- read_csv("https://data.cityofnewyork.us/resource/5uac-w243.csv?$limit=133114", na = c("(null)", ""),
                        col_types = cols(
    cmplnt_num = col_character()))
```

```{r}
complaints <- complaints |>
  mutate(complaint_date = as.Date(cmplnt_fr_dt)) |>
  rename(Precinct = addr_pct_cd) |>
  filter(complaint_date >= as.Date("2026-01-01"))

complaintsbyprecinct <- complaints |>
  count(Precinct) |>
  rename(complaints = n)

arrestsbyprecinct <- arrests |>
  count(Precinct) |>
  rename(arrests = n)

joint_data <- left_join(complaintsbyprecinct, arrestsbyprecinct, by = "Precinct")
```

```{r}
ggplot(joint_data, aes(x = complaints, y = arrests)) +
  geom_point(alpha = 0.6, stroke = 0, size = 2) +
  geom_smooth(method = "lm", color = "#FF4136", se = FALSE) +
  labs(title = "NYPD Arrests vs. NYPD Complaints by Precinct", 
       subtitle = "From January 1, 2026 to March 31, 2026") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5), plot.subtitle = element_text(hjust = 0.5))
```

The graph presents the relationship between the number of NYPD complaints and arrests across precincts, where each plotted point represents a precinct. There is a strong, positive association between complaints and arrests, suggesting that precincts with more complaints are a strong predictor of arrests ($\beta = 0.62, p < 0.001$). Finally, NYPD complaints explain 91.11% of the variation in arrests ($R^2 = 0.9111$).

```{r}
model <- lm(arrests~complaints, data = joint_data)

summary(model)
```

```{r}
shootings <- read_csv("https://data.cityofnewyork.us/resource/5ucz-vwe8.csv?$limit=25000", na = c("(null)", ""))
```

```{r}
shootings <- shootings |>
  mutate(boro = recode(boro, `MANHATTAN` = "Manhattan", `BRONX` = "Bronx", `BROOKLYN` = "Brooklyn",
                           'QUEENS' = "Queens", 'STATEN ISLAND' = "Staten Island")) |>
  rename(boro = boro, Precinct = precinct) |>
  filter(occur_date >= as.Date("2026-01-01"))
  
```

```{r}
arrests_num <- arrests |>
  count(Precinct) |>
  rename(arrests = n)

complaints_num <- complaints |>
  count(Precinct) |>
  rename(complaints = n) 

shootings_num <- shootings |>
  count(Precinct) |>
  rename(shootings = n)

data <- arrests_num |>
  full_join(complaints_num, by = "Precinct") |>
  full_join(shootings_num, by = "Precinct")

data$shootings <-  replace_na(data$shootings, 0)

```

#### What type of crime indicators are more strongly associated with NYPD arrests across NYC precincts.

```{r}
library(GGally)
ggpairs(data[, c("arrests", "complaints", "shootings")])
```
In this scatterplot matrix, each observation represents a NYC precinct. The variables, which are NYPD arrests, NYPD complaints, and NYC shootings, are summed up counts within each precinct. The pairwise plot shows relationships and their correlation coefficients between each combination of the variables, all of which are positive and statistically significant (indicated by the asterisks). The graph also suggests that precincts with higher levels of NYPD complaints are correlated with higher NYPD arrest counts, while shootings have a weaker correlation with higher arrest accounts in comparison.

### Topic 3: Racial Composition of People Arrested Across NYC

#### Does the racial distribution of arrests differ by NYC borough?

```{r}
ggplot(arrests, aes(x = arrest_boro, fill = perp_race)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  labs(y = "proportion", title = "Proportional Distribution of Arrests by Race and NYC Borough",
       fill = "Race", x = "Borough", subtitle = "From January 1, 2026 to March 31, 2026") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.2), plot.subtitle = element_text(hjust = 0.5))
```

The bar graph shows the proportional distribution of arrests by race across the five boroughs, with each borough on the x-axis. Across all boroughs, individuals identified as Black make up the largest proportion of arrests compared to other racial groups, while Asians/Pacific Islanders represent the smallest proportion


#### What is the racial distribution of people who are arrested in the precinct with the highest number of arrests?

```{r}
data$Precinct[which.max(data$arrests)]
```

```{r}
p75 <- arrests |>
  filter(Precinct == 75)

race_p75 <- p75 |>
  count(perp_race)

race_p75 <- p75 |>
  count(perp_race) |>
  mutate(prop = n / sum(n))

ggplot(race_p75, aes(y = fct_reorder(perp_race, prop), x = prop)) +
  geom_col(fill = "#3D9970") +
  theme_minimal() +
  labs(title = "Racial distribution of arrests in Precinct 75 (Brooklyn)",
       x = "Proportion", y = "Race", subtitle = "From January 1, 2026 to March 31, 2026") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))

```

As represented in the bar graph, out of the 2545 people arrested in Precinct 75, 68% were identified as Black, 18% were identified as White Hispanic, 6% were identified as Black hispanic, 4% were identified as White, 3% were identified as Asian/Pacific Islander, and less than 0% were identified as American Indian/Alaskan Native or unknown.

#### What is the racial distribution of people who are arrested in the precinct with the lowesr number of arrests?

```{r}
data$Precinct[which.min(data$arrests)]
```

```{r}
p22 <- arrests |>
  filter(Precinct == 22)

race_p22 <- p22 |>
  count(perp_race)

race_p22 <- p22 |>
  count(perp_race) |>
  mutate(prop = n / sum(n))

ggplot(race_p22, aes(y = fct_reorder(perp_race, prop), x = prop)) +
  geom_col(fill = "#3D9970") +
  theme_minimal() +
  labs(title = "Racial distribution of arrests in Precinct 22 (Manhattan)", 
       x = "Proportion", y = "Race", subtitle = "From January 1, 2026 to March 31, 2026") +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))

```

As represented in the bar graph, there were a total of 19 arrests in Precinct 22. Out of those arrests, 58% were people identified as hispanics, 32% were people identified as black, and 11% were people identified as white.


### Short Summary of Findings
Overall, there are different geographic and demographic variations in NYPD arrest patterns across NYC. Brooklyn had the highest frequency of arrests and also contained the precinct with the highest number of arrests. On the other hand, Staten Island had the lowest total number of arrests, while Manhattan's Precinct 22 recorded the lowest arrest count year to date. When comparing arrests with crime indicators, both shootings and NYPD complaints showed strong, positive correlations with arrest counts across precincts. However, the association between complaints and arrests was noticeably stronger and statistically significant, suggesting a stronger relationship between NYPD complaints and arrest activity at the precinct level. Additionally, individuals identified as Black represented the largest proportion of arrested across all five boroughs, though racial composition varied at the  precinct level.



## Conclusion

This project analyzed NYPD arrest data (YTD) to explore relationships between arrests and crime indicators as well as spatial patterns and racial composition across NYC. My project also had several limitations, including that the observed relationships are correlational and should NOT be interpreted as causal; potential influence from unobserved variables like population size, socioeconomic conditions, policing presence, income distribution, and precinct-level demographic composition; and time alignment across datasets that limited the analysis to this year and prevented a time series analysis. Future work could consider this analysis by including additional variables or sociopolitical context, particularly in relation to policing policy shifts such as stop-and-frisk. Overall, I feel like I built a lot of skills during this project and learned that even though data may not always be able to answer my questions, available data is a start to get me thinking about possible explanations. 
