# NYPD Arrest Data

By: Jade Amaya

## Introduction

Hi everyone, welcome to my project! This project looks at NYPD Arrest Data (YTD) and examines the relationships between arrests, location, demographic characteristics, and other crime indicators. I use precinct-level aggregated data to analyze how arrest patterns vary across NYC and how they relate to complaints, shootings, and race with respect to arrests. I wanted to look at this data set (and related ones) because I have a broader interest in how policing and crime intersect with historical patterns in law enforcement, specifically in the U.S. Some historical context motivating my project is related to slavery patrol systems during the antebellum period that were replaced by modern police systems during the Jim Crow era to reinforce segregation, and recently (as in a few years ago) stop-and-frisk in the late 20th to early 21st century. Amid this context, I use the following datasets (refer to Data Description) to look at statistical relationships that may tell a story as to how arrests are distributed across NYC during this time.

## Data

### Description

The data was collected by the NYPD and published by NYC’s Open Data team. The data was imported, using an csv file from the NYPD Arrest (Year to Date) Data set. The dataframe contains 19 columns and 63.5k rows. The data is updated every year some time during mid to late April. Another dataset that I use in this project is also NYPD Complaint data, using an API. To control for time, both datasets are filtered to only include data from this year (1/1/26 - 3/31/26). Some limitations of this dataset is that there is no data provided that informs us whether or not these arrested people were let go, whether they were scheduled a trial, or whether some of these arrests pertain to the same individual; that some demographic information pertaining to individuals are missing; and that even though these people were arrested, cannot say for certain all actually committed the crime.

Here are the links to the datasets:

[NYPD Arrests](https://data.cityofnewyork.us/Public-Safety/NYPD-Arrest-Data-Year-to-Date-/uip8-fykc/about_data)

[NYPD Complaints](https://data.cityofnewyork.us/Public-Safety/NYPD-Complaint-Data-Current-Year-To-Date-/5uac-w243/about_data)

### Missing Value Analysis

```{r}
library(tidyverse)
library(sf)
library(readxl)
library(patchwork)
arrests <- read_csv("data/arrests.csv", na = c("(null)", ""))
```

```{r}
arrests <- arrests |>
  mutate(mdydate = mdy(ARREST_DATE)) |>
  rename(perp_sex = PERP_SEX, age = AGE_GROUP, pd_desc = PD_DESC,
         arrest_key = ARREST_KEY, law_cat_cd = LAW_CAT_CD,
         law_code = LAW_CODE, x_coord_cd = X_COORD_CD, y_coord_cd = Y_COORD_CD,
         ky_cd = KY_CD, jurisdiction_code = JURISDICTION_CODE, 
         ofns_desc = OFNS_DESC, arrest_boro = ARREST_BORO, 
         arrest_precinct = ARREST_PRECINCT, perp_race = PERP_RACE,
         pd_cd = PD_CD, longitude = Longitude, latitude = Latitude)
```

```{r}
#| fig-width: 8
#| warning: false
library(redav)
arrests_new <- arrests |>
  select(starts_with(c("perp_sex", "age", "pd_d", "k", "o", "law_", "pd_c", "arrest_key")))
colnames(arrests_new) <- str_remove_all(colnames(arrests_new), "PE")
plot_missing(arrests_new)
```

Most of the rows in the dataset contained no missing values. Above the graph shows that the second most common pattern had sex and age group missing values. The third pattern included all other columns without missing values, but the level of defense.

## Results

### Topic 1: Location of Arrests 

#### Where are arrests concentrated?

```{r}
#|warning: false
library(sf)
nybb <- st_read("nybb_26a/nybb.shp")

nybb <- nybb |>
  mutate(BoroCode = recode(BoroCode, `1` = "M", `2` = "B", `3` = "K",
                           '4' = "Q", '5' = "S")) |>
  rename(arrest_boro = BoroCode)
```

```{r}
graph1 <- ggplot(arrests, aes(x = arrest_boro, fill = arrest_boro)) +
  geom_bar() +
  theme_minimal()
```


```{r}
#| message: false
points_sf <- arrests |> 
 select(longitude, latitude, arrest_boro) |> 
  na.omit() |> 
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326) 
graph2 <- ggplot() + 
 geom_sf(data = nybb) +
  geom_sf(data = points_sf, size = .1, aes(color = arrest_boro)) +
  theme_minimal()
```

```{r}
#| fig-length: 5
#| fig-width: 10
graph1 + graph2
```


```{r}

nypp <- read_sf("nypp_25c/nypp.shp")
arrests <- read_csv("data/arrests.csv", na = c("(null)", ""))

arrests <- arrests |>
  mutate(arrest_date = mdy(ARREST_DATE)) |>
  rename(perp_sex = PERP_SEX, age = AGE_GROUP, pd_desc = PD_DESC,
         arrest_key = ARREST_KEY, law_cat_cd = LAW_CAT_CD,
         law_code = LAW_CODE, x_coord_cd = X_COORD_CD, y_coord_cd = Y_COORD_CD,
         ky_cd = KY_CD, jurisdiction_code = JURISDICTION_CODE, 
         ofns_desc = OFNS_DESC, arrest_boro = ARREST_BORO, 
         Precinct = ARREST_PRECINCT, perp_race = PERP_RACE,
         pd_cd = PD_CD, longitude = Longitude, latitude = Latitude)

newer <- left_join(nypp, arrests) |>
  select(c("Precinct", "perp_race", "Shape_Leng", "Shape_Area", "geometry", "arrest_date")) |>
  group_by(perp_race, Precinct) |>
  summarise(race_count = n())
```

```{r}
newdata <- left_join(nypp, arrests) |>
  select(c("Precinct", "perp_sex", "Shape_Leng", "Shape_Area", "geometry")) |>
  filter(perp_sex == "M") |>
  group_by(Precinct) |>
  summarise(Male = n())

male <- ggplot(newdata) +
  geom_sf(aes(fill = Male)) + 
  geom_sf_text(aes(label = Precinct), size = 2,
               color = "white") +
   scale_fill_gradient(low = "darkblue", high = "lightblue",
                     na.value = "grey70") +
  theme_void() 
```

### Topic 2: Arrests and Indicators of Crime Across Precincts

#### What is the association between the number of crime complaints and arrests across precincts?

```{r}
complaints <- read_csv("https://data.cityofnewyork.us/resource/5uac-w243.csv?$limit=133114", na = c("(null)", ""),
                        col_types = cols(
    cmplnt_num = col_character()))
```

```{r}
complaints <- complaints |>
  mutate(complaint_date = as.Date(cmplnt_fr_dt)) |>
  rename(Precinct = addr_pct_cd) |>
  filter(complaint_date >= as.Date("2026-01-01"))

complaintsbyprecinct <- complaints |>
  count(Precinct) |>
  rename(complaints = n)

arrestsbyprecinct <- arrests |>
  count(Precinct) |>
  rename(arrests = n)

joint_data <- left_join(complaintsbyprecinct, arrestsbyprecinct, by = "Precinct")
```

```{r}
ggplot(joint_data, aes(x = complaints, y = arrests)) +
  geom_point(alpha = 0.6, stroke = 0, size = 2) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Complaints vs Arrests by Precinct") +
  theme_minimal()
```

There is a strong, positive correlation between NYPD complaints and arrests. After running a linear regression model, I found that complaints were a strong predictor for arrests($\beta = 0.62, p < 0.001$), and that NYPD complaints also explain 91.11% of the variation in arrests ($R^2 = 0.9111$). In other words, there is a strong association between NYPD complaints and arrests.

```{r}
model <- lm(arrests~complaints, data = joint_data)

summary(model)
```



```{r}
shootings <- read_csv("https://data.cityofnewyork.us/resource/5ucz-vwe8.csv?$limit=25000", na = c("(null)", ""))
```
```{r}
shootings <- shootings |>
  mutate(boro = recode(boro, `MANHATTAN` = "M", `BRONX` = "B", `BROOKLYN` = "K",
                           'QUEENS' = "Q", 'STATEN ISLAND' = "S")) |>
  rename(boro = boro, Precinct = precinct) |>
  filter(occur_date >= as.Date("2026-01-01"))
  
```

```{r}
arrests_num <- arrests |>
  count(Precinct) |>
  rename(arrests = n)

complaints_num <- complaints |>
  count(Precinct) |>
  rename(complaints = n) 

shootings_num <- shootings |>
  count(Precinct) |>
  rename(shootings = n)

data <- arrests_num |>
  full_join(complaints_num, by = "Precinct") |>
  full_join(shootings_num, by = "Precinct")

data$shootings <-  replace_na(data$shootings, 0)

```


#### What type of crime indicators are more strongly associated with NYPD arrests across NYC precincts.
```{r}
library(GGally)
ggpairs(data[, c("arrests", "complaints", "shootings")])
```
In this graph, each observation represents a NYC precinct. The variables, which are arrests, complaints, and shootings, are summed up counts within each precinct. The pairwise plot shows relationships and their correlation coefficients between each combination of the variables, all of which are positive and statistically significant (indicated by the asterisks). The graph also suggests that precincts with higher levels of NYPD complaints are associated with higher NYPD arrest counts, while shootings have a weaker association with higher arrest accounts in comparison.

### Topic 3: Arrests by Location and Race

#### Does the racial distribution of arrests differ by NYC borough?

```{r}
ggplot(arrests, aes(x = arrest_boro, fill = perp_race)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  labs(y = "proportion")
```




#### What is the racial distribution of people who are arrested in the precinct with the highest number of arrests?
```{r}
data$Precinct[which.max(data$arrests)]
```

```{r}
p75 <- arrests |>
  filter(Precinct == 75)

race_p75 <- p75 |>
  count(perp_race)

race_p75 <- p75 |>
  count(perp_race) |>
  mutate(prop = n / sum(n))

ggplot(race_p75, aes(y = fct_reorder(perp_race, prop), x = prop)) +
  geom_col(fill = "darkblue") +
  theme_minimal() +
  labs(title = "Racial distribution of arrests in Precinct 75 (Brooklyn)")

```
Out of the 2545 people arrested in Precinct 75, 68% were identified as Black, 18% were identified as White Hispanic, 6% were identified as Black hispanic, 4% were identified as White, 3% were identified as Asian/Pacific Islander, and
less than 0% were identified as American Indian/Alaskan Native or unknown.


#### What is the racial distribution of people who are arrested in the precinct with the lowesr number of arrests?
```{r}
data$Precinct[which.min(data$arrests)]
```

```{r}
p22 <- arrests |>
  filter(Precinct == 22)

race_p22 <- p22 |>
  count(perp_race)

race_p22 <- p22 |>
  count(perp_race) |>
  mutate(prop = n / sum(n))

ggplot(race_p22, aes(y = fct_reorder(perp_race, prop), x = prop)) +
  geom_col(fill = "darkblue") +
  theme_minimal() +
  labs(title = "Racial distribution of arrests in Precinct 22 (Manhattan)")

```
There were a total of 19 arrests in Precinct 22. Out of those arrests, 58%
were people identified as hispanics, 32% were people identified as black,
and 11% were people identified as white.




## Conclusion
